$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-07-17 paper/paper.tex
Take care to anonsuppress references to our mechanisation

  We present the syntax, semantics, type system and corresponding type safety properties.
  For the first time in the \picalc{}, we use leftover typing \cite{Allais2018a} to encode our typing rules in a way that propagates linearity constraints into process continuations.
  We generalise the algebras on multiplicities, allowing the user to choose a mix of linear, graded and shared typing.
  We provide framing, weakening and strengthening theorems, which can now be stated for the linear \picalc{} and use them to prove subject congruence.
  We show that the type system is stable under renaming and prove subject reduction.
%
  Our formalisation is fully mechanised in Agda \cite{Zalakain2020Agda}.
\end{abstract}

@@

    Benefits of typing with leftovers include: removing the need for extrinsic context splits, which are rendered unnecessary; and for the first time we can state theorems like \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) for the linear \picalc{}.    
  \end{itemize}
  \item \textbf{Fully mechanised formalisation and proofs of the \picalc{} with leftovers}:
  The formalisation of the \picalc{} with leftovers has been fully mechanised in Agda and is publicly available in our GitHub repository \cite{Zalakain2020Agda}.

@@

\begin{proof}[Prove (Sketch)]
  All the above theorems are proved by induction on \textsc{Process} and \textsc{Var}.
  For details, refer to our mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

\begin{proof}[Proof (Sketch)]
  All the above theorems are proved by induction on \textsc{Types} and \textsc{VarRef}.
  For details, refer to our mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  For the nontrivial base cases and their symmetric variants, we apply a combination of \autoref{thm:framing}, \autoref{thm:strengthening}, \autoref{thm:weakening} and \autoref{thm:exchange}.
  Further details can be found in \autoref{app:struct} and the full proof in \cite{Zalakain2020Agda}.
\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To conclude, in this paper we provide a well scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now we stated for the linear \picalc{}.
Our work is fully mechanised in around 1500 lines of code in Agda \cite{Zalakain2020Agda}.
%
As future work, we intend to prove further properties of our type system, such as that reduction preserves the balancing of channels.
We intend to add support for products, sum types and recursion to both our syntax and our type system.
Orthogonally, making our typing rules bidirectional would allow us to provide a decision procedure for type checking processes in a given set of algebras.
Furthermore, it might also be worth identifying correspondences between our usage algebra and particular state machines.
Finally, we will use our \picalc{} with leftovers as an underlying framework on top of which we can implement session types and other advanced type theories.

>>>>

  We present the syntax, semantics, type system and corresponding type safety properties.
  For the first time in the \picalc{}, we use leftover typing \cite{Allais2018a} to encode our typing rules in a way that propagates linearity constraints into process continuations.
  We generalise the algebras on multiplicities, allowing the user to choose a mix of linear, graded and shared typing.
  We provide framing, weakening and strengthening theorems, which can now be stated for the linear \picalc{} and use them to prove subject congruence.
  We show that the type system is stable under renaming and prove subject reduction.
%
  Our formalisation is fully mechanised in Agda.
\end{abstract}

@@

    Benefits of typing with leftovers include: removing the need for extrinsic context splits, which are rendered unnecessary; and for the first time we can state theorems like \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) for the linear \picalc{}.    
  \end{itemize}
  \item \textbf{Fully mechanised formalisation and proofs of the \picalc{} with leftovers}:
    The formalisation of the \picalc{} with leftovers has been fully mechanised in Agda and is publicly available in our GitHub repository%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.

@@

\begin{proof}[Prove (Sketch)]
  All the above theorems are proved by induction on \textsc{Process} and \textsc{Var}.
  For details, refer to our mechanisation in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

@@

\begin{proof}[Proof (Sketch)]
  All the above theorems are proved by induction on \textsc{Types} and \textsc{VarRef}.
  For details, refer to our mechanisation in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

@@

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  For the nontrivial base cases and their symmetric variants, we apply a combination of \autoref{thm:framing}, \autoref{thm:strengthening}, \autoref{thm:weakening} and \autoref{thm:exchange}.
  Further details can be found in \autoref{app:struct} and the full proof in%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To conclude, in this paper we provide a well scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now we stated for the linear \picalc{}.
Our work is fully mechanised in around 1500 lines of code in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
%
As future work, we intend to prove further properties of our type system, such as that reduction preserves the balancing of channels.
We intend to add support for products, sum types and recursion to both our syntax and our type system.
Orthogonally, making our typing rules bidirectional would allow us to provide a decision procedure for type checking processes in a given set of algebras.
Furthermore, it might also be worth identifying correspondences between our usage algebra and particular state machines.
Finally, we will use our \picalc{} with leftovers as an underlying framework on top of which we can implement session types and other advanced type theories.
